A Hedonic Price Analysis of Traffic Noise in the Twin Cities Housing Market

### 2013-06-27 reports/Draft.Rnw

\title{A Hedonic Price Analysis of Traffic Noise in the Twin Cities Housing Market}
\date{}

@@

\begin{abstract}
This paper implements the hedonic price method to estimate the impact of traffic noise on properties within the St. Paul, Minnesota metropolitan real-estate market over the time period from 2005 - 2010. We use Locally Weighted Regression (LWR) estimation techniques in order to allow the hedonic function to vary over time and space. Significant evidence suggests spatial and temporal non-stationarity in the hedonic price function within our study area. In contrast to previous research, we find little evidence that the impact of traffic noise exhibits a significant non-linear relationship with the decibel level of the traffic noise, nor does the impact seem to vary over space. Our results suggest that an increase in traffic noise of one decibel decreases house prices by an average of 0.19 percent before September, 2008, and an average of 0.37 percent after September, 2008. Both of these estimates are within the range estimated by previous research on the topic in other areas.
\end{abstract}

\section{Introduction and Past Research}\label{sec:lit}
The value of noise pollution is typically estimated through simulated or related markets because no explicit private market for noise abatement exists. \citet{Nelson1982} is perhaps the best known earliest review of 14 studies of the economic impact of traffic noise on property values in Canada and the United States. More recently, a number of reviews have been produced for European government agencies. \citet{Bateman2001} reviewed the literature for the Scottish Government while \citet{Navrud2002} provided a ``state of the art'' review for the European Commission Environment Directorate General.

@@

\citet{Nelson2008} produced an updated review of the traffic noise literature for the edited volume \underline{Hedonic Methods in Housing Markets Economics}. The chapter describes several areas of continued interest in the empirical literature, most notably: the appropriate measure of traffic noise, spatial modeling of the hedonic price function, and housing market dynamics. We now explain each of these areas in turn and our work's contribution to the field.

\subsection{Measure of Traffic Noise}

Beyond the fact that no direct natural economic markets exist in which consumers can reveal their willingness to pay to avoid traffic noise, researchers also must choose how to measure the nuisance itself. Sound levels are commonly measured in decibels (dB), but traffic noise at a house can vary significantly depending on changing traffic conditions.

Some researchers (see, for instance, \citet{HughesJr.1992}) measure the negative externality of traffic by correlating traffic volume with house prices. However, traffic may generate other externalities beyond noise (such as local air pollution and reduced safety for pedestrians and drivers). Given the interest in government policymakers in traffic noise mitigation (cite some data here), we focus on the negative externality stemming specifically from the traffic noise itself. Additionally, until 2011, the Minnesota Department of Transportation used a cost effectiveness threshold for noise mitigation of \$3,250 per dB reduced. Given that some of the benefits of noise reductions are capitalized in house prices, we are curious to see how our estimated magnitudes compare to the state cost effectiveness cutoff.

I'm going to need some help writing about noise measures. Perhaps Tsegaye can lend a hand?

\citet{Huang;Palmquist2001} collect noise data using 128 monitoring stations throughout a suburb of Seattle, WA. They use dBA $L_{10}$ as their noise measure, where $L_{10}$ denotes the value that is exceeded 10 percent of the time. They collapse the noise readings into 2.5 dBA wide bins and report noise impact estimates of roughly 1.4\% decrease in house sales price per category increase in noise (or about 0.4\% per dBA $L_{10}$).

@@

@@

How do people do submarkets? 
\begin{itemize}
  \item Geography 
    \begin{itemize}
      \item census tracts
    \end{itemize}
  \item Political Boundaries
    \begin{itemize}
      \item city boundaries
    \end{itemize}
  \item Other (Data Mining)
\end{itemize}

\citet{Theebe2004a} econometrically estimates a spatial error model

Researchers such as , and \citet{Andersson2010} attempt to correct for spatial autocorrelation through the use of spatial

Day (2003) investigated the impact of traffic noise in Glasgow, Scotland and found four separate submarkets with varying relationships between noise and prices. Interestingly, traffic noise coefficients were estimated to be negative in three out of four submarkets, while the fourth revealed a significantly positive relationship. The positive relationship is hypothesized to be related to increased transportation access and mobility in an outer suburb of the city. In a similar study, Day et. al (2007) found traffic noise estimates to be significantly negative in five of eight submarkets using appriximately 10 thousand residential sales in 1997 in Birmingham, England.

Most researchers who allow the hedonic function to vary over geography do so through explicit parameterization done by the researchers (usually using political boundaries).

Theebe (2004) also used market segmentation to estimate more than 160,000 property sale values in the western part of the Netherlands sold between 1997 and 1999.  First, the data were split into five parts according to political jurisdictions. Through spatial autoregression techniques, Theebe’s results indicated that sales values are unaffected by noise levels below 65 dB. Above 65 dB, NDSI is relatively small, ranging from 0.3 to 0.5\% per dB. An analysis of submarket data yielded larger results although these findings are less precise. While Theebe covered a large study area, the noise levels are obtained for square areas of 100 x 100m using the cumulative 24 hours energy level index Leq. This study uses a Leq index at a square area of 10 x 10m, therefore each property has a unique and continuous noise decibel measurement. Consequentially, the analysis on smaller submarket samples will be relatively more precise in terms of traffic noise estimates than Theebe (2004).

While Day (2003) and Theebe (2004) considered the existence of submarkets through market segmentation and used estimators that accounted for spatial autocorrelation, these segmentations are limited by space. By defining clear boundaries of submarkets, Duarte et. al (2009) noted that this clear delimitation does not allow interdependence between submarkets to be considered. Through locally weighted regression (LWR), the authors resolved the issue of spatial autocorrelation and considered the existence of submarkets while providing a ‘soft’ border between localities. By examining 3,196 appraisals of multifamily dwellings assessed in 2005 in Barcelona, Spain, Duarte et. al investigated a range of neighborhood and environmental attributes.  A Monte Carlo test statistically validated the spatial variation of local factors. The results from the LWR indicate that the average NDSI for observations located in the 65-70 dB is negative  while average NDSI for observations exposed to 70-75 dB is 0.26\%. Finally, dwellings exposed to higher levels of noise (75-80 dB) have an average -0.40\% NDSI. The authors explained this paradox by considering areas with intermediate levels of noise have closer access to services and transport within the city without experiencing an extreme noise externality. In this study, a similar implementation of locally weighted regression is utilized in order to resolve for spatial autocorrelation while allowing for interdependency across submarkets. Furthermore, two Monte Carlo tests are utilized to statistically support (1) the use of a non-global model and (2) the spatial variation of estimated parameters.

\subsection{Temporal Variation}
A common strategy is to analyze cross-sectional data. However, housing transactions take place at a specific point in time and the vast majority of data sets will contain some temporal variation in house sales. Some researchers may simply assume the hedonic function is constant thoughout their timespan. Others allow the function to shift up or down through the use of temporally varying intercept terms.

Few researchers allow the hedonic function to change over time. \citet{Cho2011b} is a recent example of work allowing the hedonic regression coefficients to change over time, thereby estimating the changes in the implicit prices of hedonic characteristics over time.

\citet{Wilhelmsson2000} split his 290 Swedish house sales into two separate time periods: 1986-89 and 1990-95. A Chow test revealed that the impact of traffic noise was different in the earlier vs. later time period, with noise being considered more of a nuisance in the suburb of Stockholm during the 1990s.

We estimate regression coefficients over time using a moving window of 12 months in the hedonic analysis over the years 2005 - 2010. Such an analysis allows us to see if and when the implicit price of traffic noises changes within our study area. Our results suggest that the percentage change in house price from an extra decibel of traffic noise actually became more negative after the Great Recession than before. These results are different from other hedonic work such as \citet{Cho2011b} who found that environmental amenities in the Nashville, TN area mattered less during the Great Recession compared to before it.

@@

Standard urban economic theory\footnote{Cite Muth, Mills, etc. here.} predicts that the price of land will vary over space within our dataset to account for locational ameinities. As such, we add interaction terms between lot size and distance to the nearest central business district to account for spatial variation in the price of land. The first column in Table \ref{tab:globalRegression} shows the results of this regression. The negative coefficient of -0.0027 on the noise variable (a one dB increase in noise decreases house price by 0.27 percent) is in line with the estimates described earlier in the literature. The significant coefficients on the lot size interaction terms suggest that the value of land varies over space. The significant coefficients on the non-linear interaction terms suggests that the nature of the spatial variation in the value of land may be complex.

@@

@@

We estimated the impact of traffic noise on housing prices using Locally Weighted Regression techniques in the St. Paul, MN metropolitan area. Specifically, we estimate linear regressions at each house within our dataset using only information contained in geographically nearby houses sold within the last 12 months. We find strong evidence that the hedonic function in our study area differs over space and time. Local regressions using only information contained in the nearest 200 houses generate a substantially smaller Generalized Cross Validation score (0.266) compared to the same model assuming geographical stationarity in the regression coefficients (0.401).

>>>>

\title{A Hedonic Price Analysis of Traffic Noise in the Twin Cities Housing Market}
\date{\today}

@@

\begin{abstract}
This paper implements the hedonic price method to estimate the impact of traffic noise on properties within the St.\ Paul, Minnesota metropolitan real-estate market over the time period from 2005 - 2010. We use Locally Weighted Regression (LWR) estimation techniques in order to allow the hedonic function to vary over time and space. Significant evidence suggests spatial and temporal non-stationarity in the hedonic price function within our study area. In contrast to previous research, however, we find little evidence that the impact of traffic noise exhibits a significant non-linear relationship with the decibel level of the traffic noise, nor does the impact seem to vary over space. We find significant differences in the impact of traffic noise from the beginning to the end of our sample. Our results suggest that an increase in traffic noise of one decibel decreases house prices by an average of 0.19 percent before September, 2008, and an average of 0.37 percent after September, 2008. While significantly different, both of these estimates are still within the typical range of previous estimates in the literature.
\end{abstract}

\section{Introduction and Past Research}\label{sec:lit}
The value of noise pollution is typically estimated through simulated or related markets because no explicit private market for noise abatement exists. \citet{Nelson1982} is perhaps the best known earliest review of 14 housing market studies of the economic impact of traffic noise on property values in Canada and the United States. More recently, a number of reviews have been produced for European government agencies. \citet{Bateman2001} reviewed the literature for the Scottish Government while \citet{Navrud2002} provided a ``state of the art'' review for the European Commission Environment Directorate General.

@@

\citet{Nelson2008} produced an updated review of the traffic noise literature for the edited volume \underline{Hedonic Methods in Housing Markets Economics}. The chapter describes several areas of continued interest in the empirical literature, most notably: the measure of traffic noise, spatial modeling of the hedonic price function, and housing market dynamics.

\subsection{Measure of Traffic Noise}
I'm going to need some help writing about noise measures. Perhaps Tsegaye can lend a hand?

Noise data is typically not collected for each every house in a hedonic study. For example, \citet{Huang;Palmquist2001} collect noise data using 128 monitoring stations throughout a suburb of Seattle, WA. Such a strategy forces the researchers to interpolate the data between the monitoring stations in order to associate each house with a noise level. While the researchers do not describe how the interpolation was done, they implicitly admit the potential uncertainty of the method by categorizing traffic noise exposure into 2.5 dB wide categories. Other researchers, such as \citet{Theebe2004a} use 5 dB wide bins, and \citet{Nelson2008} even claims that some studies use bins as wide as 10 dB. Such coarse measurements can lead to serious errors in variables and imprecise estimates of the impact of noise.

Other researchers obtain noise levels from external sources without much description of the variable source. For instance, \cite{MarmolejoDuarteCarlos;GonzalezTamez2009} simply list a ``City Council Sonic Map'',  \citet{Blanco2011} cite the Department for Environment Food and Rural Affairs Noise mapping England project. \citet{Andersson2010} state, ``Information about noise levels is from a study on the health effects of traffic noise conducted in Lerum [Sweden] in 2004 (Ohrstrom et al. 2005). Separate noise calculations were made for railway and road noise for all the houses in Lerum.''

Traffic noise at a location can vary significantly depending on changing traffic conditions. Such variation has led to multiple measures of noise. Some measures eliminate the fluctuations in sound and instead represent the equivalent continuous constant sound pressure at a site. Other indices use noise values that are only exceeded a given percentage of time. For instance, \citet{Huang;Palmquist2001} use dB $L_{10}$ as their noise measure, where $L_{10}$ denotes the value that is exceeded only 10 percent of the time. Still others adjust nighttime noise exposure levels to account for potential increased perceived nuisance. \citet{Baranzini2010} even surveyed inhabitants about their perceptions of the level of noise and their annoyance. The work finds similar results for significant levels of measured noise as compared subjective measures of noise.

\emph{What do we have? There is still a lot of work to be done in this area, better understanding the measures others use as well as what we have and how they compare. I'm surprised at how little information people actually provide about their noise data.}

Spatial resolution of the data. \citet{Theebe2004a} had noise levels for square areas of 100 x 100m.

% Beyond the fact that no direct natural economic markets exist in which consumers can reveal their willingness to pay to avoid traffic noise, researchers also must choose how to measure the nuisance itself. Some researchers (see, for instance, \citet{HughesJr.1992}, \citet{Li2012} \citet{Larsen2012}) attempt to measure the negative externality of traffic by correlating traffic volume with house prices. However, traffic may generate other externalities beyond noise (such as local air pollution and reduced safety for pedestrians and drivers) and so these strategies overestimate the size of the noise externality. 
% 
% The value of traffic noise is important in and of itself. Governments routinely erect noise barriers to shield residential areas from traffic noise. (cite some numbers here!). To aid such work, it is important to understand the benefits of these policies. For instance, until 2011, the Minnesota Department of Transportation used a cost effectiveness threshold for noise mitigation projects of \$3,250 per dB reduced per house. Given that some of the benefits of noise reductions are capitalized in house prices, we are curious to see how our estimated magnitudes compare to the state cost effectiveness cutoff.

@@

It is now common to use spatial econometric techniques in the hedonic literature, for instance estimating a spatial lag (in which house prices are modeled as a function of neighboring house prices), or a spatial error model (in which the degree of spatial autocorrelation is also estimated). In the noise literature, for example, \citet{Theebe2004a} econometrically estimates a spatial error model, while \citet{Andersson2010} uses a spatial lag model.

@@

Some traffic noise research simply explicitly assumes that their data comprise a single market. For example, \citet{Kim2007} assumes that by restricting analysis to residential properties surrounding the Inner Circular highway in the central area of Seoul from 2002 to 2004, that the price of traffic noise will be constant. Other research, such as \citet{Theebe2004a} used political boundaries to segment the market into five separate subarkets in the western part of the Netherlands.

Another strategy is to let the data reveal different relationships within smaller areas of the study. For instance, \citet{Day2007} and \citet{MarmolejoDuarteCarlos;GonzalezTamez2009} use semi-parametric techniques to allow the hedonic relationship to smoothly vary over space. We use techniques similar to the Geographically Weighted Regression method described in \citet{MarmolejoDuarteCarlos;GonzalezTamez2009} to allow the hedonic function to vary over space without having a priori conceptions of where the function will be different.

\subsection{Temporal Variation}
Little work has been done estimating the changes in hedonic environmental (dis)amenities like noise over time. An event-study literature exists using temporal variation in noise, especially noise from airports (see, for example, \citet{Cohen2009}, ). However, these articles usually use the change in noise over time to identify the hedonic price of noise, rather than ask whether the hedonic price has changed.

\citet{Wilhelmsson2000} is one example of research asking whether the implicit price of noise has changed over time. The study splits the 290 Swedish house sales into two separate time periods: 1986-89 and 1990-95. A Chow test revealed that the impact of traffic noise was different in the earlier vs. later time period, with noise being considered more of a nuisance in the suburb of Stockholm during the 1990s.

\citet{Cho2011b} is a recent example of work asking whether the hedonic regression coefficients change over time in the Nashville, TN area. In particular, the study uses a repeat sales methodology to estimate changes in the implicit prices of hedonic characteristics as a result of the Great Recession of 2008-09 in the United States. The work finds that environmental amenities matter less in sales during the year 2008 when compared to pre-recession sales.

We estimate regression coefficients over time using a moving window of 12 months in the hedonic analysis over the years 2005 - 2010. Such an analysis allows us to see if and when the implicit price of traffic noise changes within our study timeframe. Our results suggest that the percentage change in house price from an extra decibel of traffic noise actually became more negative after the Great Recession than before.

Our work makes three contributions to the field. First, we utilize new estimates of traffic noise that has never before been used in the hedonic literature. The data explicitly model noise propagation as a function of nearby buildings and landcover, as well as present the results at a fine spatial resolution (10m x 10m). Such estimates, combined with sptially explicit modeling techniques allow us to make our second contribution to the literature. We present one of the first estimates of how the impact of traffic noiseeee varies over space. We find that while the impact of other important control variables in the hedonic function vary over space, the impact of traffic noise does not. Lastly, we present the first estimates of how the Great Recession has affected the hedonic price of traffic noise.

@@

Standard urban economic theory\footnote{Cite Muth, Mills, etc. here.} predicts that the price of land will vary over space within our dataset to account for locational amenities. As such, we add interaction terms between lot size and distance to the nearest central business district to account for spatial variation in the price of land. The first column in Table \ref{tab:globalRegression} shows the results of this regression. The negative coefficient of -0.0027 on the noise variable (a one dB increase in noise decreases house price by 0.27 percent) is in line with the estimates described earlier in the literature. The significant coefficients on the lot size interaction terms suggest that the value of land varies over space. The significant coefficients on the non-linear interaction terms suggests that the nature of the spatial variation in the value of land may be complex.

@@

We didn't explicitly model spatial autocorrelation in the error terms.

We might want to further reduce the number of parameters to estimate through the use of either mixed-LWR or running separate regressions for submarkets determined through hierarchical clustering like \citet{Day2003}.

@@

We estimated the impact of traffic noise on housing prices using Locally Weighted Regression techniques in the St.\ Paul, MN metropolitan area. Specifically, we estimate linear regressions at each house within our dataset using only information contained in geographically nearby houses sold within the last 12 months. We find strong evidence that the hedonic function in our study area differs over space and time. Local regressions using only information contained in the nearest 200 houses generate a substantially smaller Generalized Cross Validation score (0.266) compared to the same model assuming geographical stationarity in the regression coefficients (0.401).
